---
title: "Getting started with clintlf"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with clintlf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(clintlf)
```

**clintlf** is a small, dependency-free collection of formatting and data
utilities for producing tables, listings, and figures (TLFs). Whether you work
in base R, the tidyverse, or a proprietary pipeline, it integrates into
existing workflows without tying you to a specific framework, package
ecosystem, or dependency stack.

This vignette walks through three functional areas:

## 1. Table formatting

### Half-away-from-zero rounding

Base R rounds half to even (banker's rounding), while most clinical reporting
conventions use half-away-from-zero rounding. `fmt_round()` applies this
behavior and returns a character vector with trailing zeros preserved, ready
for table display.

```{r}
round(c(9.5, 10.5), 0)      # base R: half to even
fmt_round(c(9.5, 10.5), 0)  # half away from zero

fmt_round(c(2.5, 0.035, NA), c(0, 2, 2))
```

Rounding and display are one step here: you specify decimal places, not
significant figures, and get back a string that already reads the way the table
should.

### Descriptive statistics cells

`fmt_stat()` builds the usual summary cells from a lead value and a bracketed
group: mean (SD), median (Q1, Q3), an estimate with its bounds, or min, max with
no lead. The lead and the bracketed part round independently. The `na` text
stands in for a missing value, like the SD of a single observation.

```{r}
fmt_stat(23.456, 4.5678, inner_n = 2)                    # mean (SD)
fmt_stat(3.14159, 1.23456, 5.6789, n = 2, inner_n = 2)   # estimate (95% CI)
fmt_stat(NULL, 0.123, 9.876)                             # min, max (no lead, no brackets)
fmt_stat(1, 0.5, 2.5, bracket = "(]")                    # inclusive/exclusive range
fmt_stat(c(12.3, 15.8), c(4.56, NA), na = "-")
```

### Counts with percentages

`fmt_percent()` formats `"xx (xx%)"`. If a positive count rounds to 0%, it
switches to a `"<"` threshold label so the cell doesn't read as zero when the
count isn't. A zero denominator leaves the percentage undefined, so those cells
print `zero_denom` (default `"-"`).

```{r}
fmt_percent(15, 100)
fmt_percent(1, 1000, perc_n = 0)
fmt_percent(1, 1000, perc_n = 1)
fmt_percent(0, 0)
```

### P-values

`fmt_pval()` displays values below a significance threshold as `"<alpha"` and
rounds the rest half away from zero. Pass the threshold as a character string
to preserve trailing zeros in the label.

```{r}
fmt_pval(c(0.051, 0.001), n = 3, alpha = "0.05")
```

## 2. Date imputation

`impute_date()` fills the missing month and day of partial date strings, the
kind that turn up in international clinical data. It handles `"ymd"`, `"mdy"`,
and `"dmy"` orderings, auto-detects the separator, and strips any time component
first. By default a missing month fills to `"12"` and a missing day to the last
day of the resolved month (the end-of-period convention); pass
`fill_month = "01"`, `fill_day = "01"` for the earliest-date convention.

```{r}
impute_date(c("2025-07-19", "2025-07", "2025", NA))
impute_date(c("2025-07-19", "2025-07"), fill_month = "01", fill_day = "01")
impute_date("2026-05-22T15:30:45")
impute_date("07-19-2025", fmt = "mdy")
```

Out-of-range month or day components in the output trigger a warning, but
calendar validity is not checked (custom fill values can intentionally create
dates such as `"2025-02-31"`). Validate with `as.Date()` when calendar-valid
dates are required.

## 3. Data dictionary extraction

`dict_param()` summarizes unique `PARAM` / `PARAMCD` / `PARAMN` combinations
from ADaM datasets; `dict_label()` extracts variable label attributes as
carried by CDISC datasets (e.g., read via `haven::read_xpt()`). Both accept a
single data frame or a named list of data frames.

```{r}
adlb <- data.frame(
  USUBJID = c("001", "001", "002", "002"),
  PARAMCD = c("ALB", "ALB", "BILI", "BILI"),
  PARAM   = c("Albumin (g/L)", "Albumin (g/L)",
              "Bilirubin (umol/L)", "Bilirubin (umol/L)"),
  PARAMN  = c(1L, 1L, 2L, 2L),
  AVAL    = c(4.1, 4.2, 0.8, 0.9)
)

dict_param(adlb)
```

```{r}
attr(adlb$USUBJID, "label") <- "Unique Subject Identifier"
attr(adlb$AVAL,    "label") <- "Analysis Value"

dict_label(adlb)
```
