---
title: "Getting started with clintlf"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with clintlf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(clintlf)
```

**clintlf** is a lightweight, standalone collection of formatting and data
utilities for producing tables, listings, and figures (TLFs). Whether you work
in base R, the tidyverse, or a proprietary pipeline, it integrates into
existing workflows without tying you to a specific framework, package
ecosystem, or dependency stack.

This vignette walks through four functional areas:

## 1. Table formatting

### Half-away-from-zero rounding

Base R rounds half to even (banker's rounding). Many clinical reporting
conventions, as well as SAS `ROUND()`, use half-away-from-zero rounding.
`fmt_round()` applies this behavior and returns a character vector with trailing
zeros preserved, ready for table display.

```{r}
round(c(9.5, 10.5), 0)      # base R: half to even
fmt_round(c(9.5, 10.5), 0)  # half away from zero

fmt_round(c(2.5, 0.035, NA), c(0, 2, 2))
```

The function separates numerical rounding from display formatting, allowing
numeric results to follow reporting conventions based on decimal-place
specifications rather than significant figures.

### Estimates with confidence intervals

`fmt_ci()` combines a point estimate and its bounds into `"xx (xx, xx)"`,
with independent decimal control for the estimate and the bounds.

```{r}
fmt_ci(3.14159, 1.23456, 5.67890)
fmt_ci(3.14159, 1.23456, 5.67890, est_n = 2, ci_n = 1)
```

### Counts with percentages

`fmt_percent()` formats `"xx (xx%)"`. When a positive count produces a
percentage that rounds to zero, it substitutes a `"<"` threshold label so the
table never claims a zero percentage for a non-zero count. A zero denominator
prints `zero_denom` (default `"-"`) in place of the undefined percentage.

```{r}
fmt_percent(15, 100)
fmt_percent(1, 1000, perc_n = 0)
fmt_percent(1, 1000, perc_n = 1)
fmt_percent(0, 0)
```

### P-values

`fmt_pval()` displays values below a significance threshold as `"<alpha"` and
rounds the rest half away from zero. Pass the threshold as a character string
to preserve trailing zeros in the label.

```{r}
fmt_pval(c(0.051, 0.001), n = 3, alpha = "0.05")
```

## 2. Date imputation

`impute_date()` imputes missing month and day components of partial date
strings with varying structures, commonly encountered in international clinical
datasets. It supports `"ymd"`, `"mdy"`, and `"dmy"` orderings, auto-detects
separators, and strips time components before imputation.

```{r}
impute_date(c("2025-07-19", "2025-07", "2025", NA))
impute_date(c("2025-07-19", "2025-07"), fill_month = "12", fill_day = "31")
impute_date("2026-05-22T15:30:45")
impute_date("07-19-2025", fmt = "mdy")
```

Out-of-range month or day components in the output trigger a warning, but
calendar validity is not checked (custom fill values can intentionally create
dates such as `"2025-02-31"`). Validate with `as.Date()` when calendar-valid
dates are required.

## 3. Data dictionary extraction

`dict_param()` summarizes unique `PARAM` / `PARAMCD` / `PARAMN` combinations
from ADaM datasets; `dict_label()` extracts variable label attributes as
carried by CDISC datasets (e.g., read via `haven::read_xpt()`). Both accept a
single data frame or a named list of data frames.

```{r}
adlb <- data.frame(
  USUBJID = c("001", "001", "002", "002"),
  PARAMCD = c("ALB", "ALB", "BILI", "BILI"),
  PARAM   = c("Albumin (g/L)", "Albumin (g/L)",
              "Bilirubin (umol/L)", "Bilirubin (umol/L)"),
  PARAMN  = c(1L, 1L, 2L, 2L),
  AVAL    = c(4.1, 4.2, 0.8, 0.9)
)

dict_param(adlb)
```

```{r}
attr(adlb$USUBJID, "label") <- "Unique Subject Identifier"
attr(adlb$AVAL,    "label") <- "Analysis Value"

dict_label(adlb)
```

## 4. Multiple imputation pooling

`calc_poolmi()` pools estimates and standard errors across imputed datasets
using Rubin's rules, without requiring adoption of an entire multiple imputation
framework. Supply the per-imputation estimates and standard errors directly as
vectors, or as columns of a stacked data frame (one row per imputation).

Two degrees-of-freedom methods are available: `df = "1987"` (Rubin, 1987;
approximates the SAS `PROC MIANALYZE` default) and `df = "1999"` (Barnard and
Rubin, 1999; approximates `mice::pool()` and `PROC MIANALYZE` with EDF).

```{r}
calc_poolmi(est = c(-2.57, -2.11, -1.85, -2.07, -2.13),
            se  = c( 1.05,  0.91,  0.92,  0.85,  0.90))
```

The same pooled results are obtained from a stacked analysis-results data
frame:

```{r}
imp_result <- data.frame(
  IMPNO = 1:5,
  est   = c(-2.57, -2.11, -1.85, -2.07, -2.13),
  SE    = c( 1.05,  0.91,  0.92,  0.85,  0.90)
)

calc_poolmi(est = "est", se = "SE", data = imp_result)
```

### From a stacked imputed dataset

In practice, the per-imputation results come from fitting the same model in
each imputed dataset. Clinical MI datasets are typically stacked, with an
`IMPNO` column identifying the imputation. To keep this example runnable, the
"imputations" below are simulated by adding noise to the missing outcomes:

```{r}
set.seed(18210)
subj <- data.frame(
  USUBJID = sprintf("SUBJ-%03d", 1:60),
  TRTP    = factor(rep(c("Placebo", "Drug"), each = 30),
                   levels = c("Placebo", "Drug")),
  BASE    = round(rnorm(60, 25, 4), 1)
)
subj$CHG <- -1 - 2 * (subj$TRTP == "Drug") + rnorm(60, 0, 3)
miss <- sample(60, 12)   # subjects with a missing outcome

admi <- do.call(rbind, lapply(1:5, function(m) {
  imp <- subj
  imp$CHG[miss] <- imp$CHG[miss] + rnorm(length(miss), 0, 2)
  cbind(IMPNO = m, imp)
}))
```

Split by imputation, fit a baseline-adjusted ANCOVA in each, and stack the
treatment effect and its standard error — one row per imputation:

```{r}
admi_list <- split(admi, admi$IMPNO)

model_results <- do.call(rbind, lapply(admi_list, function(dat) {
  coefs <- summary(lm(CHG ~ BASE + TRTP, data = dat))$coefficients
  data.frame(est = coefs["TRTPDrug", "Estimate"],
             SE  = coefs["TRTPDrug", "Std. Error"])
}))

calc_poolmi(est = "est", se = "SE", data = model_results)
```

Any model — mixed models included — slots into the same pattern, as long as
each imputation-specific fit supplies an estimate and its standard error.

### Cross-check against mice

`calc_poolmi()` with `df = "1999"` reproduces `mice::pool()`:

```{r, eval = requireNamespace("mice", quietly = TRUE)}
imp <- mice::mice(mice::nhanes, m = 5, print = FALSE, seed = 18210)
fit <- with(data = imp, lm(bmi ~ age))

summary(mice::pool(fit), conf.int = TRUE)[2, ]

est_se <- do.call(rbind, lapply(fit$analyses, function(model) {
  coefs <- summary(model)$coefficients
  data.frame(est = coefs["age", "Estimate"],
             SE  = coefs["age", "Std. Error"])
}))

calc_poolmi(est = "est", se = "SE", data = est_se,
            df = "1999", n = nrow(mice::nhanes), k = 2)
```

The equivalent SAS calls, given the same stacked dataset:

```
PROC MIANALYZE DATA = est_se;            /* df = "1987" */
  MODELEFFECTS est;
  STDERR SE;
RUN;

PROC MIANALYZE DATA = est_se EDF = 23;   /* df = "1999", EDF = n - k */
  MODELEFFECTS est;
  STDERR SE;
RUN;
```

## References

- Rubin, D.B. (1987). *Multiple Imputation for Nonresponse in Surveys*. Wiley.
- Barnard, J. and Rubin, D.B. (1999). Small-sample degrees of freedom with
  multiple imputation. *Biometrika*, 86(4), 948-955.
